PyUltraLight: A Pseudo-Spectral Solver for Ultralight Dark Matter Dynamics

### 2018-06-14 draft.tex
Changes made in response to Shaun's comments. A few notes added.

\abstract{\PyUltraLight  simulates the dynamics of ultralight dark matter in a static, non-expanding \sh{are static and non-expanding not exactly rhe same?} background.  \PyUltraLight can describe the evolution of several interacting ultralight dark matter halos  or one or more halos orbiting a central, fixed Newtonian potential, the latter scenario corresponding to dwarf galaxies orbiting a massive central galaxy. We verify  \PyUltraLight  by showing that it reproduces qualitative dynamical features of previously published simulations and demonstrate that it has  excellent energy-conservation properties.   \PyUltraLight is implemented in a Python-based Jupyter notebook, solving the Schr{\"o}dinger-Poisson equation  governing  ultralight scalar field dark matter dynamics in the non-relativistic regime using a symmetrised split-step pseudospectral  algorithm. The notebook interface makes it simple to specify simulation parameters and  visualise the resulting output but performance-critical routines are managed via calls to computationally efficient compiled libraries. \PyUltraLight  runs on standard desktop hardware with support for shared memory mutlithreading and is available on GitHub.  \re{small change}}

@@

The potential shortcomings of simple cold dark matter scenarios motivate investigations of more novel dark matter scenarios. In particular, ultralight dark matter (ULDM), also known as fuzzy dark matter (FDM), or BEC dark matter, is an increasingly well-studied possibility; for a recent review of the potential advantages and characteristic attributes of this scenario see Ref~\cite{Hui2016}.   ULDM models are well motivated by fundamental theories possessing approximate shift symmetries\sh{suggest include ", e.g. an axion" with references. Otherwise this sentence seems a bit out of place, hinting at something without being explicit.}. Moreover, ULDM can naturally resolve the small-scale problems of $\Lambda$CDM as the Heisenberg uncertainty principle suppresses gravitational collapse on length scales shorter than the de Broglie wavelength of the ULDM particle. In this regime the mass of the ULDM particle becomes correlated with astrophysical observables; if it is on the order of $10^{-22}$~eV, structure is suppressed at kiloparsec scales and below \cite{Hu2000}.

@@

The existence of an extremely light scalar field, minimally coupled to gravity, is the central premise on which ULDM models are predicated. Within the ULDM framework, it is proposed that cosmological structure formation proceeds through the Bose-Einstein condensation of this scalar field \sh{I don't understand this sentence. There is a Bose-Einstein condensate and it clusters to form structures, but the sentence seems to suggest that the clustering and condensating are the same thing?}. In this Section  \re{I would normally give Section a capital letter in this context} \ek{I would be inclined to only capitalise `section' in instances where it appears as a proper noun, i.e. "In Section X we showed...", where we are referring to a specifically named section, numbered X. It seems to me that in this particular context `section' is a common noun. Happy to capitalise all instances of the word if this is your recommendation.} 
\re{I agree, but thought ``this Section'' was an example of it being used as a proper noun. But it is not a huge deal either way.}\ek{The way I would tend to think of it is by analogy with the following: "This person studies physics." vs. "Emily studies physics." In the former sentence, "person" is used as a description rather than a title, so would not be capitalised. Similar to "Richard is a professor." vs. "This is Professor Richard."}
we derive the (non-relativistic) dynamical equations of the single macroscopic condensate wavefunction, namely, the Schr{\"o}dinger-Poisson equations. We begin with the action functional for a scalar field, $\phi$, minimally coupled to gravity and in the absence of self-interactions,

@@

It is helpful to recast the Schr{\"o}dinger-Poisson system (equations \ref{eq:schrodinger} and \ref{eq:poisson}) in terms of adimensional quantities. In keeping with Refs~\cite{Schive2014,Paredes2016} we introduce length, time, and mass scales as follows:
\begin{align}
    \CMcal{L}&=\left(\frac{8\pi\hbar^2}{3 m^2H_0^2\Omega_{m_0}}\right)^{\frac{1}{4}}\approx121\left(\frac{10^{-23}\operatorname{eV}}{m}\right)^{\frac{1}{2}}\operatorname{kpc},\label{eq:length}\\
    \CMcal{T}&=\left(\frac{8\pi}{3 H_0^2\Omega_{m_0}}\right)^{\frac{1}{2}}\approx75.5 \operatorname{Gyr},\label{eq:time}\\
    \CMcal{M}&=\frac{1}{G}\left(\frac{8\pi}{3 H_0^2\Omega_{m_0}}\right)^{-\frac{1}{4}}\left(\frac{\hbar}{m}\right)^{\frac{3}{2}}\approx 7\times 10^7\left(\frac{10^{-23}\operatorname{eV}}{m}\right)\operatorname{M}_{\odot},\label{eq:mass}
\end{align}
where $m$ is the mass of the ultralight scalar field, $H_0$ is the present-day Hubble parameter, $G$ is Newton's gravitational constant and $\Omega_{m_0}$ is the present-day matter fraction of the energy density of the universe. We  recast equations \ref{eq:schrodinger} and \ref{eq:poisson} in terms of the  dimensionless quantities
\begin{equation}\label{eq:dimensionless}
    t'=\frac{t}{\CMcal{T}},\quad
    \vec{x}^{\,'}=\frac{\vec{x}}{\CMcal{L}},\quad
    \Phi'=\frac{m\CMcal{T}}{\hbar}\Phi,\quad
    \psi'=\CMcal{T}\sqrt{mG}\psi.
\end{equation}
Dropping the primes for notational convenience, we see that the coupled differential equations of the  Schr{\"o}dinger-Poisson system  reduce to 
\begin{align}
    i\Dot{\psi}(\vec{x},t)&=-\frac{1}{2}\nabla^2\psi(\vec{x},t)+\Phi(\vec{x},t)\psi(\vec{x},t),\label{eq:s-adim}\\
    \nabla^2\Phi(\vec{x},t)&=4\pi\vert\psi(\vec{x},t)\vert^2,\label{eq:p-adim}
\end{align}
where it is understood that all quantities involved are dimensionless. We can recover  dimensionful quantities via the ``dictionary'' provided by equations \ref{eq:length} to \ref{eq:mass}. For example, the integrated mass of the system, $M_{tot}$, is given by
\begin{equation}\label{eq:integrated-mass}
    M_{tot}=\CMcal{M}\int d^3x\vert\psi\vert^2.
\end{equation}
Likewise, the mass density at any point is given by
\begin{equation}\label{eq:density}
    \rho=\frac{\CMcal{M}}{\CMcal{L}^3}\vert\psi\vert^2.
\end{equation}
\PyUltraLight\ works internally with these dimensionless quantities but receives initial conditions and generates output in physical units. Henceforth, we will often refer to $\vert\psi\vert^2$  as the density, where it is understood that this is in fact a dimensionless quantity related to the physical mass density via the constant of proportionality given by equation \ref{eq:density}.

@@

In place of the CFL condition, the \PyUltraLight timestep can be set via the fluid interpretation of the Schr{\"o}dinger-Poisson system. In dimensionless code units we can write 
\begin{equation}
    \psi\equiv\sqrt{\rho}e^{i\theta}, \quad \vec{v}\equiv\boldsymbol{\nabla}\theta.
\end{equation}
We define a `maximum velocity' for the simulation grid at which point the fluid appears to move backward because the phase change between successive grid points exceeds $\pi$. An upper bound can then be placed on the timestep so that fluid travelling at this maximum  velocity traverse a distance of one grid space per timestep.  In practice this can yield a very demanding constraint on the timestep and for many scenarios the actual velocity is far less than this hypothetical limit. Consequently, the timestep can be  significantly increased at the user's discretion by changing the `step\textunderscore factor' parameter in the preliminary section of the Jupyter notebook. When this is done, convergence testing on a case-by-case basis is strongly encouraged.\sh{I wonder whether it is worth mentioning that the two pieces of the split step method are exact to arbitrary large times when the other piece isn't present.}\ek{I'm not quite sure what you mean by 'when the other piece isn't present', but I have a feeling that bringing this up here might cause confusion}

@@

\begin{figure}
  \includegraphics[width=1.\textwidth,trim=1cm 1cm 0 1cm,clip]{riemann}
  \caption{Solitonic core after one revolution around a central potential, demonstrating the internal velocity field of an irrotational Riemann-S ellipsoid (where the collective motion of the soliton has been subtracted from the velocity field). Deformation of the spherical density profile into that of an ellipsoid can be seen, with extension along the radial direction (red arrow). The host:satellite mass ratio is approximately 55; simulation resolution is $256^3$. \sh{I assume the contours are density, but this is never explicitly stated.}}
  \label{fig:riemann}
\end{figure}

@@

An extensive study of the tidal disruption of ULDM solitonic cores orbiting a central potential has recently been undertaken in \cite{Du2018} and we reproduce just one of their results here. When the Schr{\"o}dinger-Poisson system is recast as a system of hydrodynamical equations, the fluid velocity  at any point is the gradient of the smoothly-varying phase\sh{This correspondence is only approximate though, right? So the thing we call 'fluid velocity' must have zero curl, but the actual movement of mass doesn't necessarily need to be?}.  Since the curl of a gradient vanishes identically, the velocity field of the Schr{\"o}dinger-Poisson system must be irrotational, and $\nabla\times\vec{v}=0$. If a radially symmetric soliton is initialised in a circular orbit around a Newtonian potential, there will be initial transient behaviour as the spherical profile becomes elongated along the radial direction of the central potential. Meanwhile, the velocity field corresponding to the overall orbital motion of the soliton is superposed with the internal velocity field, combining so as to produce a net flow with vanishing curl.

The family of Riemann-S ellipsoids describe uniformly rotating bodies which satisfy the condition of irrotationality \cite{Chandrasekhar1965} \sh{this sentence is really confusing, but I'm not sure of how to re-phrase it to make it less confusing so we might need to just accept the "rotating bodies... that are irrotational" confusingness}. Therefore we expect the characteristic internal velocity field of such an ellipsoid to arise during our simulations \sh{I split the previous sentence in two and re-ordered what is now the second sentence}. It is found in \cite{Du2018} that an initially spherical solitonic core without self-rotation will gradually spin up to form a tidally-locked irrotational Riemann-S ellipsoid when orbiting a host mass. We reproduce this result using \PyUltraLight. Figure \ref{fig:riemann} showing the internal velocity field of a solitonic satellite after one complete revolution around a host mass \sh{I spluit one sentence into two}. The soliton has become elongated along the radial line connecting it to the host, indicating that it is tidally locked, while the internal velocity field qualitatively matches that of an irrotational Riemann-S ellipsoid within the tidal radius \cite{Daller2012} \sh{This is stated as if the reader should be able to tell, but I would guess most of our readers haven't heard of a Riemann-S ellipsoid. I would suggest at least saying "see figure 2 of ref [37]", and perhaps even some more substantial description of hwat one expects a riemann s ellipsoid to look like. alternatively, we could not use that jargon and just appeal to the fact that the velocity field in our figure doesn't look rotational.}.

It should be noted that the wider velocity field is not expected to be accurately predicted in a simulation of this kind. This is because the initial soliton density profile is defined only out to a given cutoff radius, beyond which the $\psi$ field value, and by extension its velocity field, is set identically to zero. Therefore, the velocity field far from the solitonic centre is not well-modelled in these simulations, and it may be expected that non-trivial self-interaction in the wider field would lead to unpredictable behaviour in these regions. \sh{This last paragraph was previously part of the preceding one. I found it a bit confusing because it reads like a defence of an anomaly in our own figure. I was going to suggest removing it but I think I now understand that it was put in as a warning to people who might use our code and get confused about the velocity field far from the centre of the solitons. We could either make that more clear or just hope that making it a separate paragraph does this implicitly}

%\vspace{1em}

@@

\begin{figure}
  \includegraphics[width=1.1\textwidth,trim=2.5cm 0 0 1cm,clip]{combined_energy_and_density1}
  \caption{Left: Evolution of the density profile of a solitonic core at equally spaced times as it undergoes tidal disruption in a potential centred at the red cross. Right: Evolution of the energy of the system, where times are indicated in correspondence with the snapshots of the density profile. All quantities are in dimensionless code units.\sh{It might be worth repeating here what the energy code units are because it is stated in the text but not in the main 'code units' section.}}
  \label{fig:combined_1}
\end{figure}

\begin{figure}
  \includegraphics[width=0.9\textwidth,trim=0 0 2cm 2cm,clip]{egy_m=22.eps}
  \caption{Evolution of the energy for a binary soliton system with each soliton in an elliptical orbit around the common centre of mass.\sh{Should probably add what energy code units are here too as people might not know to look at other figure}}
  \label{fig:binary}
\end{figure}

\begin{figure}
  \includegraphics[width=0.9\textwidth,trim=0 0.5cm 0 0,clip]{Total_Energy_Change.eps}
  \caption{Demonstration of improving energy conservation with simulation resolution for a binary system of solitons orbiting their common centre of mass.}
  \label{fig:energy_change}
\end{figure}

\begin{figure}
  \includegraphics[width=0.9\textwidth,trim=0 0.5cm 0 0,clip]{Total_Energy_Change_Disruption.eps}
  \caption{Demonstration of improving energy conservation with simulation resolution for a soliton undergoing significant tidal disruption in a Newtonian central potential.}
  \label{fig:energy_change_2}
\end{figure}

@@

The two terms in the integral \ref{eq:energy-tot} are calculated separately within the code. The first term is the gravitational potential energy of the Schr{\"o}dinger-Poisson system, $E_{GP}$ \sh{I added $E_{GP}$ here because it is labelled like this in the figures and this makes it easier for people looking at the figures to find out what that quantity is. same for $E_K$ and $E_Q$ below}. As discussed in \cite{Hui2016}, the second term may be decomposed into contributions which may be considered separately as kinetic and `quantum' energies, $E_K$ and $E_Q$. However, for our purposes it is sufficient to consider only their combined contribution\sh{split one sentence into three}. When \PyUltraLight includes the central potential of a point mass located at the centre of the simulation grid we have additional energy contributions and calculate the  gravitational potential energy from self-interactions separately from the gravitational potential energy due to the central potential.

@@

Figure \ref{fig:binary} demonstrates the evolution of the energy of a binary system of solitons in elliptical orbits around their common centre of mass over three orbital periods. In dimensionless code units, the soliton masses are 22, the initial separation  2, and the initial relative velocity was 3.6. At points of closest approach the kinetic energy increases as the solitons speed up, while the potential energy due to self-interaction decreases commensurately such that the total energy is conserved. In this scenario no central potential has been included. As the solitons reach the first point of closest approach, they become slightly deformed, exciting oscillatory modes which are manifest in the Figure as small scale oscillations superposed on the global behaviour. Figure \ref{fig:energy_change} demonstrates the relationship between the total integrated energy and the grid resolution for the same binary system of solitons used to generate Figure~\ref{fig:binary} \sh{I think people can probably work out what the y-axis is in this figure and figure 7, but it isn't ever explicitly stated. It might be good to state it here and/or in the figure captions}. Energy is conserved at sub-percent level even at low resolutions ($96^3$), and increasing grid resolution greatly improves accuracy.

@@

The current implementation of \PyUltraLight is already a useful tool for simulating dynamical ULDM systems and exploring their dynamics However, there is much scope for  improvement. In particular, future releases may incorporate a variable timestep and more sophisticated physics, including explicit self-interactions in the axion sector or additional matter components\sh{We could also mention the inclusion of other types of central potentials}.  \PyUltraLight is publicly available under a BSD license.

@@

\PyUltraLight is publicly available under a BSD licence. The full repository, including supplementary files such as the code used to generate soliton profiles, is available on GitHub at {\url{ https://github.com/erckendall/PyUltraLight}}. \PyUltraLight makes use of the pyFFTW pythonic wrapper around the FFTW C-based fast Fourier transform libraries. Both pyFFTW and FFTW are freely-available and \PyUltraLight has been used successfully on both Mac OS and Linux  systems\sh{I've also used it on the cluster quite easily}. We welcome advice and feedback from users.

>>>>

\abstract{\PyUltraLight  simulates the dynamics of ultralight dark matter in a non-expanding background. \PyUltraLight can describe the evolution of several interacting ultralight dark matter halos  or one or more halos orbiting a central, fixed Newtonian potential, the latter scenario corresponding to dwarf galaxies orbiting a massive central galaxy. We verify  \PyUltraLight  by showing that it reproduces qualitative dynamical features of previously published simulations and demonstrate that it has  excellent energy-conservation properties.   \PyUltraLight is implemented in a Python-based Jupyter notebook, solving the Schr{\"o}dinger-Poisson equation  governing  ultralight scalar field dark matter dynamics in the non-relativistic regime using a symmetrised split-step pseudospectral  algorithm. The notebook interface makes it simple to specify simulation parameters and  visualise the resulting output but performance-critical routines are managed via calls to computationally efficient compiled libraries. \PyUltraLight  runs on standard desktop hardware with support for shared memory mutlithreading and is available on GitHub.  \re{small change}}

@@

The potential shortcomings of simple cold dark matter scenarios motivate investigations of more novel dark matter scenarios. In particular, ultralight dark matter (ULDM), also known as fuzzy dark matter (FDM), or BEC dark matter, is an increasingly well-studied possibility; for a recent review of the potential advantages and characteristic attributes of this scenario see Ref~\cite{Hui2016}. ULDM models are well motivated by fundamental theories possessing approximate shift symmetries such as the theory of the QCD axion \cite{Kim2010, Marsh2016}. Moreover, ULDM can naturally resolve the small-scale problems of $\Lambda$CDM as the Heisenberg uncertainty principle suppresses gravitational collapse on length scales shorter than the de Broglie wavelength of the ULDM particle. In this regime the mass of the ULDM particle becomes correlated with astrophysical observables; if it is on the order of $10^{-22}$~eV, structure is suppressed at kiloparsec scales and below \cite{Hu2000}.

@@

The existence of an extremely light scalar field, minimally coupled to gravity, is the central premise on which ULDM models are predicated. Within the ULDM framework, it is proposed that this scalar field exists as a Bose-Einstein condensate, described by a single wavefunction which is governed by the Schr{\"o}dinger-Poisson coupled differential equations. We begin by deriving this system of equations as a non-relativistic weak-field limit of a more general theory. We start from the action functional for a scalar field, $\phi$, minimally coupled to gravity and in the absence of self-interactions,

@@

It is helpful to recast the Schr{\"o}dinger-Poisson system (equations \ref{eq:schrodinger} and \ref{eq:poisson}) in terms of adimensional quantities. In keeping with Refs~\cite{Schive2014,Paredes2016} we introduce length, time, and mass scales as follows:
\begin{align}
    \CMcal{L}&=\left(\frac{8\pi\hbar^2}{3 m^2H_0^2\Omega_{m_0}}\right)^{\frac{1}{4}}\approx121\left(\frac{10^{-23}\operatorname{eV}}{m}\right)^{\frac{1}{2}}\operatorname{kpc},\label{eq:length}\\
    \CMcal{T}&=\left(\frac{8\pi}{3 H_0^2\Omega_{m_0}}\right)^{\frac{1}{2}}\approx75.5 \operatorname{Gyr},\label{eq:time}\\
    \CMcal{M}&=\frac{1}{G}\left(\frac{8\pi}{3 H_0^2\Omega_{m_0}}\right)^{-\frac{1}{4}}\left(\frac{\hbar}{m}\right)^{\frac{3}{2}}\approx 7\times 10^7\left(\frac{10^{-23}\operatorname{eV}}{m}\right)\operatorname{M}_{\odot},\label{eq:mass}
\end{align}
where $m$ is the mass of the ultralight scalar field, $H_0$ is the present-day Hubble parameter, $G$ is Newton's gravitational constant and $\Omega_{m_0}$ is the present-day matter fraction of the energy density of the universe. We  recast equations \ref{eq:schrodinger} and \ref{eq:poisson} in terms of the  dimensionless quantities
\begin{equation}\label{eq:dimensionless}
    t'=\frac{t}{\CMcal{T}},\quad
    \vec{x}^{\,'}=\frac{\vec{x}}{\CMcal{L}},\quad
    \Phi'=\frac{m\CMcal{T}}{\hbar}\Phi,\quad
    \psi'=\CMcal{T}\sqrt{mG}\psi.
\end{equation}
Dropping the primes for notational convenience, we see that the coupled differential equations of the  Schr{\"o}dinger-Poisson system  reduce to 
\begin{align}
    i\Dot{\psi}(\vec{x},t)&=-\frac{1}{2}\nabla^2\psi(\vec{x},t)+\Phi(\vec{x},t)\psi(\vec{x},t),\label{eq:s-adim}\\
    \nabla^2\Phi(\vec{x},t)&=4\pi\vert\psi(\vec{x},t)\vert^2,\label{eq:p-adim}
\end{align}
where it is understood that all quantities involved are dimensionless. We can recover  dimensionful quantities via the ``dictionary'' provided by equations \ref{eq:length} to \ref{eq:mass}. For example, the integrated mass of the system, $M_{tot}$, is given by
\begin{equation}\label{eq:integrated-mass}
    M_{tot}=\CMcal{M}\int d^3x\vert\psi\vert^2.
\end{equation}
Likewise, the mass density at any point is given by
\begin{equation}\label{eq:density}
    \rho=\CMcal{M}\CMcal{L}^{-3}\vert\psi\vert^2.
\end{equation}
By dimensional analysis, we can easily restore dimensionful units to any of the quantities calculated by the code. In particular, in the following sections it is to be understood that
\begin{equation}\label{eq:density}
    E=\CMcal{M}\CMcal{L}^2\CMcal{T}^{-2}\ E_{code}, \quad v=\CMcal{L}\CMcal{T}^{-1}\ v_{code}, 
\end{equation}
where $E$ and $v$ represent energy and velocity, respectively. \PyUltraLight\ works internally with these dimensionless quantities but can receive initial conditions and generates output in physical units. Henceforth, we will often refer to $\vert\psi\vert^2$  as the density, where it is understood that this is in fact a dimensionless quantity related to the physical mass density via the constant of proportionality given by equation \ref{eq:density}.

@@

In place of the CFL condition, the \PyUltraLight timestep can be set via the fluid interpretation of the Schr{\"o}dinger-Poisson system, discussed in greater detail in Section \ref{sec:disruption}. In dimensionless code units we define 
\begin{equation}\label{eq:madelung}
    \psi\equiv\sqrt{\rho}e^{i\theta}, \quad \vec{v}\equiv\boldsymbol{\nabla}\theta.
\end{equation}
We define a `maximum velocity' for the simulation grid at which point the fluid appears to move backward because the phase change between successive grid points exceeds $\pi$. An upper bound can then be placed on the timestep so that fluid travelling at this maximum  velocity traverse a distance of one grid space per timestep.  In practice this can yield a very demanding constraint on the timestep and for many scenarios the actual velocity is far less than this hypothetical limit. Consequently, the timestep can be  significantly increased at the user's discretion by changing the `step\textunderscore factor' parameter in the preliminary section of the Jupyter notebook. When this is done, convergence testing on a case-by-case basis is strongly encouraged.\sh{I wonder whether it is worth mentioning that the two pieces of the split step method are exact to arbitrary large times when the other piece isn't present.}\ek{I'm not quite sure what you mean by `when the other piece isn't present', but I have a feeling that bringing this up here might cause confusion}

@@

\begin{figure}
  \includegraphics[width=1.\textwidth,trim=1cm 1cm 0 1cm,clip]{riemann}
  \caption{Contours of constant density are shown for a solitonic core after one revolution around a central potential. The contours are superimposed upon a background representing the internal velocity field (where bulk motion has been subtracted). The velocity field is qualitatively that of an irrotational Riemann-S ellipsoid, and deformation of the density profile of the soliton along the radial direction (red arrow) can be seen. The host:satellite mass ratio is approximately 55; simulation resolution is $256^3$.}
  \label{fig:riemann}
\end{figure}

@@

An extensive study of the tidal disruption of ULDM solitonic cores orbiting a central potential has recently been undertaken in \cite{Du2018} and we reproduce just one of their results here. To do this, we again adopt the definitions \ref{eq:madelung}, namely
\begin{equation}\label{eq:madelung_repeated}
    \psi\equiv\sqrt{\rho}e^{i\theta}, \quad \vec{v}\equiv\boldsymbol{\nabla}\theta,
\end{equation}
where we are working in dimensionless code units. Using these definitions, the Schr{\"o}dinger-Poisson system can be recast in terms of hydrodynamical quantities in the so-called Madelung representation. The Madelung equations resemble the continuity and Euler equations of classical fluid dynamics, with the addition of a `quantum pressure' term accounting for resistance against gravitational collapse. The Madelung formalism is discussed in detail in \cite{Suarez2011, Suarez2015, Johnston2010, Kopp2017}. Because this hydrodynamical formulation defines the fluid velocity as the gradient of the phase of the field $\psi$, problems arise when $\psi=0$, where the phase is not well defined. Because of this issue, the Madelung and Schr{\"o}dinger representations are not strictly equivalent unless a quantisation condition is imposed, as discussed in \cite{Wallstrom1994}. We do not consider the subtleties of the Madelung representation here, as it is sufficient for our purposes to consider the fluid velocity in the region of a solitonic core, where no field nodes are present \footnote{It should be noted that, restoring dimensionful units, the fluid velocity $\vec{v}$ is related to the ususal quantum mechanical probability current $\vec{j}$ through $\vert\psi\vert^2\vec{v} = \vec{j} = \hbar/2mi\left[\psi^*\nabla\psi-\psi\nabla\psi^*\right]$}. For a discussion of the possible remedies to the `nodes problem', the reader is referred to Chapter 15.3 of \cite{Wyatt2005}.\sh{This correspondence is only approximate though, right? So the thing we call 'fluid velocity' must have zero curl, but the actual movement of mass doesn't necessarily need to be?} \ek{The equivalence between the Madelung representation and the SP system is exact in the regime where $\psi\neq 0$. In other words, we don't introduce any new approximations to get to the hydrodynamical representation, we just make a definition of a `fluid velocity'. So for a system that is governed by the SP equations, it is true to say that the field of these velocity vectors is strictly irrotational. If you look at my footnote, I've also written how the fluid velocity is related to the quantum mechanical probability current (or in this case, number density current since $\vert\psi\vert^2=n$). Therefore, if you multiply the fluid velocity by the mass density you do get the mass current. So yes, the irrotationality applies to the motion of the mass as well. In saying that, the SP system itself is only a non-relativistic weak field approximation anyway, so the correspondence of either representation to the fundamental physics is not exact in that sense.}. 
Where the Madelung representation is well defined, i.e. where the phase is a smoothly varying function, the velocity field of the Schr{\"o}dinger-Poisson system is strictly irrotational, $\nabla\times\vec{v}=0$. If a radially symmetric soliton is initialised in a circular orbit around a Newtonian potential, there will be initial transient behaviour as the spherical profile becomes elongated along the radial direction of the central potential. Meanwhile, the velocity field corresponding to the overall orbital motion of the soliton will be superposed with the internal velocity field, combining so as to produce a net flow with vanishing curl.

The family of Riemann-S ellipsoids describe non-axisymmetric uniformly rotating bodies whose internal velocity fields have vanishing curl \cite{Chandrasekhar1965}. \ek{Rephrased from previous sentence, removed the word `irrotational' - does this read ok?}. Therefore, it is the characteristic internal velocity field of a Riemann-S ellipsoid which we expect to arise during our simulation of a soliton orbiting a central mass. It is found in \cite{Du2018} that an initially spherical solitonic core without self-rotation will gradually spin up to form a tidally-locked ellipsoid with an irrotational internal velocity field when orbiting a host mass. We reproduce this result using \PyUltraLight. Figure \ref{fig:riemann} shows the internal velocity field of a solitonic satellite after one complete revolution around a host mass. The soliton has become elongated along the radial line connecting it to the host, indicating that it is tidally locked, while the velocity field within the tidal radius is visibly irrotational and bears the qualitative trademarks of the Riemann-S ellipsoid as presented in Figure 2 of \cite{Daller2012}.

It should be noted that the wider velocity field is not expected to be accurately predicted in a simulation of this kind, though the field within the tidal radius is well-modelled. This is because the initial soliton density profile is defined only out to a given cutoff radius, beyond which the $\psi$ field value is set identically to zero. As mentioned previously, the Madelung hydrodynamical formulation of the Schr{\"o}dinger-Poisson system is not valid where $\psi=0$. Because of this, we focus primarily on the internal velocity field within the high density region of the solitonic core. As we have seen, in this region \PyUltraLight is able to accurately reproduce the expected velocity field characteristics.\ek{Rephrased this paragraph}

@@

\begin{figure}
  \includegraphics[width=1.1\textwidth,trim=2.5cm 0 0 1cm,clip]{combined_energy_and_density1}
  \caption{Left: Evolution of the density profile of a solitonic core at equally spaced times as it undergoes tidal disruption in a potential centred at the red cross. Right: Evolution of the energy of the system, where times are indicated in correspondence with the snapshots of the density profile. All quantities are in dimensionless code units.\sh{It might be worth repeating here what the energy code units are because it is stated in the text but not in the main 'code units' section.}\ek{Now this is explicitly written in the code units section}}
  \label{fig:combined_1}
\end{figure}

\begin{figure}
  \includegraphics[width=0.9\textwidth,trim=0 0 2cm 2cm,clip]{egy_m=22.eps}
  \caption{Evolution of the energy for a binary soliton system with each soliton in an elliptical orbit around the common centre of mass.}
  \label{fig:binary}
\end{figure}

\begin{figure}
  \includegraphics[width=0.9\textwidth,trim=0 0.5cm 0 0,clip]{Total_Energy_Change.eps}
  \caption{Demonstration of improving energy conservation with simulation resolution for a binary system of solitons orbiting their common centre of mass. $\Delta E_{total}$ is the difference between the current integrated energy at a given timestep and the initial integrated energy for the simulation. The ratio of this difference to the initial integrated energy is plotted on the y axis for different resolutions.\ek{More explicit description of y axis added}}
  \label{fig:energy_change}
\end{figure}

\begin{figure}
  \includegraphics[width=0.9\textwidth,trim=0 0.5cm 0 0,clip]{Total_Energy_Change_Disruption.eps}
  \caption{Demonstration of improving energy conservation with simulation resolution for a soliton undergoing significant tidal disruption in a Newtonian central potential. Again, $\Delta E_{total}$ is the difference between the current integrated energy at a given timestep and the initial integrated energy for the simulation. The ratio of this difference to the initial integrated energy is plotted on the y axis for each resolution.}
  \label{fig:energy_change_2}
\end{figure}

@@

The two terms in the integral \ref{eq:energy-tot} are calculated separately within the code. The first term is the gravitational potential energy of the Schr{\"o}dinger-Poisson system, $E_{GP}$. As discussed in \cite{Hui2016}, the second term may be decomposed into contributions which may be considered separately as kinetic and `quantum' energies, $E_K$ and $E_Q$. However, for our purposes it is sufficient to consider only their combined contribution. When \PyUltraLight includes the central potential of a point mass located at the centre of the simulation grid we have additional energy contributions and calculate the  gravitational potential energy from self-interactions separately from the gravitational potential energy due to the central potential.

@@

Figure \ref{fig:binary} demonstrates the evolution of the energy of a binary system of solitons in elliptical orbits around their common centre of mass over three orbital periods. In dimensionless code units, the soliton masses are 22, the initial separation  2, and the initial relative velocity was 3.6. At points of closest approach the kinetic energy increases as the solitons speed up, while the potential energy due to self-interaction decreases commensurately such that the total energy is conserved. In this scenario no central potential has been included. As the solitons reach the first point of closest approach, they become slightly deformed, exciting oscillatory modes which are manifest in the Figure as small scale oscillations superposed on the global behaviour. Figure \ref{fig:energy_change} demonstrates the relationship between the total integrated energy and the grid resolution for the same binary system of solitons used to generate Figure~\ref{fig:binary}. The vertical axis shows the ratio of the deviation in the total energy to the initial value of the energy, where the deviation is measured as the difference between the current and initial values. Energy is conserved at sub-percent level even at low resolutions ($96^3$), and increasing grid resolution greatly improves accuracy.

@@

The current implementation of \PyUltraLight is already a useful tool for simulating dynamical ULDM systems and exploring their dynamics However, there is much scope for  improvement. In particular, future releases may incorporate a variable timestep and more sophisticated physics, including explicit self-interactions in the axion sector or additional matter components\sh{We could also mention the inclusion of other types of central potentials}\ek{I had this in here originally but Richard's recommendation was that we leave it out as we don't want to `reveal our hand'}.  \PyUltraLight is publicly available under a BSD license.

@@

\PyUltraLight is publicly available under a BSD licence. The full repository, including supplementary files such as the code used to generate soliton profiles, is available on GitHub at {\url{ https://github.com/erckendall/PyUltraLight}}. \PyUltraLight makes use of the pyFFTW pythonic wrapper around the FFTW C-based fast Fourier transform libraries. Both pyFFTW and FFTW are freely-available and \PyUltraLight has been used successfully on both Mac OS and Linux systems, as well as a shared-memory cluster environment. We welcome advice and feedback from users.
